Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-17 Manuscript/manuscript.tex
Edit to: abstract, intro

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the best idealization of the ion binding as direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
The new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

Direct measurements of ion-membrane interactions in real biological systems are difficult. Simplified lipid bilayers are often used instead as entry-level models to shed light on the role of ions in complex biological membranes~\cite{scherer87,seelig90,cevc90}. For this reason, interactions of biologically relevant cations, especially Na$^+$ and Ca$^{2+}$, with zwitterionic phosphocholine (PC) bilayers have been widely studied in experiments~\cite{akutsu81, altenbach84, seelig90, cevc90, tocanne90, binder02, pabst07, uhrikova08} and classical MD simulations~\cite{bockmann03, bockmann04, Berkowitz12, melcrova16, javanainen17}. The details of ion binding are, however, not fully consistent in the literature. Non-invasive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering, and infrared spectroscopies mainly suggest that Na$^+$ ions exhibit negligible binding to PC lipid bilayers. On the contrary, Ca$^{2+}$ is observed to specifically bind to a couple of PC molecules using their phosphate groups~\cite{hauser76, hauser78, herbette84, akutsu81, altenbach84, binder02, pabst07, uhrikova08}. On the other hand, most atomistic resolution molecular dynamics (MD) simulation models predict a stronger binding for the cations~\cite{catte16}. For example, simulations report Na$^+$ accumulating in the lipid interface~\cite{bockmann03} or Ca$^{2+}$ binding up to 4 PC lipids simultaneously, including not only interactions with phosphates but also with carbonyl oxygens~\cite{bockmann04, melcrova16, javanainen17}.
%{\color{red} Also, several experiments have also been interpreted to support the predictions from earlier MD simulations~\cite{bockmann03, vacha09a}. \todo{remove? Why we need this?}}

Recent studies within the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16} made an attempt to resolve these apparent controversies. A direct comparison of ion binding affinities to PC bilayers between simulations and experiments was performed using the electrometer concept~\cite{seelig87}. Namely, the changes in NMR order parameters of the headgroup upon addition of ions is directly compared to the MD simulations results. Analyzing massive amounts of data collected by the Open Collaboration method, it was concluded that the accuracy of the current state of the art lipid models for MD simulations is not sufficient for a detailed interpretation of the interactions of cations with PC lipid bilayers~\cite{catte16}.

@@

In this work, we improve the cation binding of a POPC bilayer in MD simulations via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. For this, we apply a simple mean-field electronic continuum correction (ECC)~\cite{leontyev11} which implies scaling the partial charges of the atoms to account for the missing electronic polarizability in MD simulations. Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. The newly developed ECC-POPC model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while improving the membrane binding affinities significantly to sodium and calcium cations.

>>>>

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the \todo{Joe: I'm not sure about this: best idealization} of the ion binding as direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
The new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

Direct measurements of ion-membrane interactions in real biological systems are difficult. 
Hence, simplified lipid bilayers are often used as entry-level models to shed light on the role of ions in complex biological membranes~\cite{scherer87,seelig90,cevc90}. 
For this reason, interactions of biologically relevant cations, especially Na$^+$ and Ca$^{2+}$, with zwitterionic phosphocholine (PC) bilayers have been widely studied in experiments~\cite{akutsu81, altenbach84, seelig90, cevc90, tocanne90, binder02, pabst07, uhrikova08} and classical MD simulations~\cite{bockmann03, bockmann04, Berkowitz12, melcrova16, javanainen17}. The details of ion binding are, however, not fully consistent in the literature. Non-invasive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering, and infrared spectroscopies mainly suggest that Na$^+$ ions exhibit negligible binding to PC lipid bilayers. On the contrary, Ca$^{2+}$ is observed to specifically bind to a couple of PC molecules using their phosphate groups~\cite{hauser76, hauser78, herbette84, akutsu81, altenbach84, binder02, pabst07, uhrikova08}. On the other hand, most atomistic resolution molecular dynamics (MD) simulation models predict a stronger binding for the cations~\cite{catte16}. For example, simulations report Na$^+$ accumulating in the lipid interface~\cite{bockmann03} or Ca$^{2+}$ binding up to 4 PC lipids simultaneously, including not only interactions with phosphates but also with carbonyl oxygens~\cite{bockmann04, melcrova16, javanainen17}.
%{\color{red} Also, several experiments have also been interpreted to support the predictions from earlier MD simulations~\cite{bockmann03, vacha09a}. \todo{remove? Why we need this?}}

Recent studies within the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16} made an attempt to resolve these apparent controversies. A direct comparison of ion binding affinities to PC bilayers between simulations and experiments was performed using the electrometer concept~\cite{seelig87}. Namely, the changes in NMR order parameters of the headgroup upon addition of ions is directly compared to the MD simulations results. 
Analyzing massive amounts of data collected by an open collaboration method, 
it was concluded that the accuracy of the current state of the art lipid models for MD simulations is not sufficient for a detailed interpretation of the interactions of cations with PC lipid bilayers~\cite{catte16}.

@@

In this work, we improve the cation binding in a POPC bilayer in MD simulations 
via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. 
For this, we apply an electronic continuum correction (ECC)~\cite{leontyev11}, 
in which scaling of the atomic partial charges is used to account for the missing electronic polarizability in MD simulations. 
Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. 
The newly developed ECC-POPC model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while significantly improving the membrane binding affinities to sodium and calcium cations.
